---
title: "Stein's method"
author:
  - Ian Hamilton
  - Stefan Stein
  - David Selby
date: June 2021
output:
  html_document:
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, echo = TRUE)
```

Libraries.

```{r}
library(tidyverse)
library(igraph)
```

Rescale odds so that the probabilities add to one.

```{r}
df <- read_csv("Group odds 10Jun (inc Group).csv") %>%
  rename("odds_1" = "1", "odds_draw" = "X", "odds_2" = "2") %>%
  mutate(p_ij = 1 / odds_1,
         p_draw = 1 / odds_draw,
         p_ji = 1/ odds_2,
         total_prob = p_ij + p_draw + p_ji) %>%
  mutate(p_ij = p_ij / total_prob,
         p_draw = p_draw / total_prob,
         p_ji = p_ji / total_prob) %>%
  select(-total_prob) %>%
  mutate(logit_p_ij = log(p_ij / p_ji))

df
```

Create adjacency matrix.

```{r}
makeX_lm <- function(df){
  X_lm <- matrix(0,
                 nrow(df),
                 length(levels(factor(as.character(df$TeamName1)))))
  colnames(X_lm) <- levels(factor(as.character(df$TeamName1)))
  for (team in colnames(X_lm)) {
    X_lm[df$TeamName1 == team, team] <- 1
    X_lm[df$TeamName2 == team, team] <- -1
  }
  return(X_lm)
}

df$X <- makeX_lm(df)
```

Fit linear model.

```{r}
lmfit_group <- lm(logit_p_ij ~ -1 + X, data = df)

r <- coef(lmfit_group)
r[is.na(r)] <- 0
r
s <- exp(r)
```

Scale the Group probabilities.

Under Bradley--Terry, the probability of $i$ winning the final, given $i$ has made it to the final, is
\[
q_{io} = \pi_i / (\pi_i + \pi_o)
\]
where $\pi_i$ is the strength of team $i$ and $\pi_o$ is the strength of the final opponent.
The opponent strength $\pi_o$ is assumed to be the same for all teams (since there are so many permutations of reaching the final), so
\begin{equation}
\text{logit}(q_{io}) = \lambda_i - \lambda_o
\end{equation}
where $\lambda_i = \log \pi_i$.
We have the intra-group strengths $s_i$, but we do not know the relative strengths of the groups.
Suppose group $G$ has strength $\gamma_G$ such that $s_i = \gamma_G(i) \pi_i$ where $G(i)$ is $i$'s group.
Then
\begin{equation}
\lambda_i = \log \pi_i = \log s_i - \log \gamma_{G(i)} = r_i - \lambda_{G(i)}
\end{equation}
where $\lambda_{G(i)} = \log \gamma_{G(i)}$.
So combining (1) and (2) we have
\[
r_i - \text{logit}(q_io) = \lambda_{G(i)} + \lambda_o.
\]
So we estimate $\lambda_o$ and the $\lambda_G$ by regressing against $r_i - \text{logit}(q_{io})$.

```{r}
# take tournament win odds and translate to tournament win odds


winner <- read_csv("Winner.csv")%>%
  mutate(p_i = 1 / Odds,
         total_prob = sum(p_i))%>%
  mutate(p_i = p_i / total_prob)%>%
  select(Group, TeamName, p_i)%>%
  rename("pi_win" = "p_i")%>%
  left_join(
    read_csv("making_final.csv")%>%
      mutate(p_i = 1 / Odds,
             total_prob = sum(p_i))%>%
      mutate(p_i = 2*p_i / total_prob)%>%
      select(TeamName, p_i)%>%
      rename("pi_final" = "p_i"),
    by = "TeamName"
  )%>%
  # calculate the conditional probability of winning given you make the final
  mutate(q_io = pi_win / pi_final)%>%
  left_join(
    as.data.frame(r)%>%
      rownames_to_column("TeamName")%>%
      mutate(TeamName = str_remove(TeamName, "^X")),
    by = "TeamName"
  )


```

```{r}

y <- winner$r - qlogis(winner$q_io)

X <- matrix(0,
               nrow(winner),
               length(levels(factor(as.character(winner$Group)))))
colnames(X) <- levels(factor(as.character(winner$Group)))
rownames(X) <- winner$TeamName
for (group in colnames(X)) {
  X[winner$Group == group, group] <- 1
}
X

lmfit_win <- lm(y ~ X)

lambda_G <- coef(lmfit_win)
lambda_G[is.na(lambda_G)] <- 0
lambda_G
names(lambda_G) <- c("O", "A", "B", "C", "D", "E", "F")
```

Get the team coefficients $\lambda_i = r_i - \lambda_{G(i)}$

```{r}
lambda <- setNames(winner$r - lambda_G[winner$Group], winner$TeamName)
lambda
```

Then calculate the probabilities from the $\lambda$s

```{r}
probs <- plogis(outer(lambda, lambda, '-'))
probs
```


